---
title: "STA 113 2.0 Descriptive Statistics"
subtitle: "Measures of Central Tendency and Variability: Grouped Data"
author: "Dr. Thiyanga S. Talagala <br/>  Department of Statistics, Faculty of Applied Sciences <br/> University of Sri Jayewardenepura, Sri Lanka"
format:
  revealjs:
    width: 2000
    height: 2000
    margin: 0.1
    theme: [simple, custom.scss]
    transition: slide
    background-transition: fade
    slide-number: true
    show-slide-number: all
    title-slide-attributes: 
      data-background-color: "#081d58"
      data-background-image: none
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tsibble)
library(lubridate)
library(feasts)
library(denguedatahub)
# install.packages("devtools")
#devtools::install_github("thiyangt/TourSriLanka")
library(TourSriLanka)
```

## Data

```{r}
c1 <- c("(1, 3]", "(3, 5]", "(5, 7]",
        "(7, 9]", "(9, 11]", "(11, 13]")
f <- c(16, 2, 4, 3, 9, 6)

df <- data.frame(c1, f)
df |>
  knitr::kable( col.names = c("Class interval",
                           "Frequency (f)"))

```

## Data

```{r}
c1 <- c("(1, 3]", "(3, 5]", "(5, 7]",
        "(7, 9]", "(9, 11]", "(11, 13]")
f <- c(16, 2, 4, 3, 9, 6)
m <- c(2, 4, 6, 8, 10, 12)

df <- data.frame(c1, f, m)
df |>
  knitr::kable( col.names = c("Class interval",
                           "Frequency (f)",
                           "Class midpoint (M)"))

```

---

# Mean

$$Mean = \frac{\sum fM}{\sum f}$$

---

# Mode

In-class demo

Plot the histogram.

---

# Median

$$Median = L_M + [\frac{\frac{(\sum f)}{2}-F}{f_M}]W$$

$L_M$ - lower limit of the class containing the median

$\sum f$ - total number of observations

$F$ - the cumulative frequency up to the lower limit of the class containing the median

$f_M$ - frequency of the class containing the median

$W$ - width of the class containing the median

---

# Quartiles

$$Q_1 = L_{q1} + [\frac{\frac{(\sum f)}{4}-F_{q1}}{f_{q1}}]W$$

$L_{q1}$ - lower limit of the class containing the first quartile

$\sum f$ - total number of observations

$F_{q1}$ - the cumulative frequency up to the lower limit of the class containing the first quartile

$f_M$ - frequency of the class containing the median

$W$ - width of the class containing the median

> Your turn: Write the formula for Q3


# Population variance

$$\sigma^2 = \frac{\sum f(M-\mu)^2}{\sum f}$$

$M$ - Class midpoint


# Sample variance

$$s^2 = \frac{\sum f(M-\bar{x})^2}{\sum f -1}$$

$M$ - Class midpoint