---
title: "STA 113 2.0 Descriptive Statistics"
subtitle: "Spearman's rank correlation"
author: "Dr. Thiyanga S. Talagala <br/>  Department of Statistics, Faculty of Applied Sciences <br/> University of Sri Jayewardenepura, Sri Lanka"
format:
  revealjs:
    width: 2000
    height: 2000
    margin: 0.1
    theme: [simple, custom.scss]
    transition: slide
    background-transition: fade
    slide-number: true
    show-slide-number: all
    title-slide-attributes: 
      data-background-color: "#081d58"
      data-background-image: none
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tsibble)
library(lubridate)
library(feasts)
library(denguedatahub)
# install.packages("devtools")
#devtools::install_github("thiyangt/TourSriLanka")
library(TourSriLanka)
```

## Spearman's rank correlation

-   Pearson's correlation coefficient: at least interval level of measurement for the data

-   Spearman's rank correlation: at least ordinal-level or ranked data

$r = 1 - \frac{6\sum_{i=1}^nd^2}{n(n^2-1)}$

where

$n = \text{number of pairs being correlated.}$

$d = \text{the difference in the ranks of each pair}$

------------------------------------------------------------------------

## Example

Compute Spearman's rank correlation for the following variables to determine the degree of association between the two variables.

```{r}
x <- c(10, 10, 4, 20, 30, 40, 40, 30, 20, 10)

y <- c(3, 3, 1, 2, 3, 4.5, 5, 6, 7, 1)
df <- data.frame(x=x, y=y)
df |> knitr::kable()
```

# Important

-   Pearson's correlation determines the strength and direction of the linear relationship between two variables.

-   Spearman's correlation determines the strength and direction of the monotonic relationship.

-   The Spearman's rank correlation formula is derived from the Pearson product moment formula and utilises the ranks of the $n$ pairs instead of the raw data.

------------------------------------------------------------------------

## In-class demonstration

-   Monotonic vs Non-monotonic

-   Monotonic & Non-linear vs Monotonic & Linear
