---
title: "STA 113 2.0 Descriptive Statistics"
subtitle: "Summary Measures (Cont.)"
author: "Dr. Thiyanga S. Talagala <br/>  Department of Statistics, Faculty of Applied Sciences <br/> University of Sri Jayewardenepura, Sri Lanka"
format:
  revealjs:
    width: 2000
    height: 2000
    margin: 0.1
    theme: [simple, custom.scss]
    transition: slide
    background-transition: fade
    slide-number: true
    show-slide-number: all
    title-slide-attributes: 
      data-background-color: "#081d58"
      data-background-image: none
---




## Variance of a sample

The variance of a sample of `n` observations $x_1, x_2, x_3,..,x_n$ having mean $\bar{x}$ is defined as 

$$s^2 = \frac{\sum_{i=1}^n(x_i - \bar{x})^2}{n-1}$$

Alternatively, the sample variance can be written in the following forms:

$$s^2 = \frac{1}{n-1}[\sum_{i=1}^n x_i^2 - \frac{(\sum_{i=1}^nx_i)^2}{n}]$$

or 

$$s^2 = \frac{1}{n-1}[\sum_{i=1}^nx_i^2-n\bar{x}]$$
---

## Measures of relative standing/ Measures of noncentral location

- Quartiles

- Percentiles

---

## Quantiles

Quantiles are descriptive measures that split the ordered data into four quarters (four equal parts).

Q1 - first (lower) quantile

Q2 - second (middle) quantile

Q3 - third (upper) quantile

> First quantile

The value which 25% of the observations are smaller and 75% are larger 

$$Q_1 = \frac{n+1}{4} \text{ ordered observation}$$

> Second quantile

Same as median

> Third quantile 

The value for which 75% of the observations are smaller and 25% are larger

$$Q_3 = \frac{3(n+1)}{4} \text{ ordered observation}$$


## Rules

1. If the resulting positioning point is an integer, take the particular value corresponding to that positioning point.

2. For a non-integer position 
$p$, let $k$ be the integer part and $d$ be the fractional part (e.g., for 
$p=2.75$, $k=2$ and $d=0.75$)

$$Q_q = (1-d) \text{ value at position }k + d.\text{ value at position }k+1$$

> Your turn

Find quantiles for 

1,3,4,6,7,8,10,12,14,15

## Percentiles

- Percentiles divides a given ordered data array into 100 equal parts, it divides the complete data set into hundred groups of 1% each. There are total of 99 percentiles denoted as P1, P2, P3,…, and, P99, and they are known as 1st percentile, 2nd percentile,…., 99th percentile respectively.

first decile = 10th percentile

Q1 = 25th percentile

Q2 = 50th percentile

Q3 = 75th percentile

ninth decile = 90th percentile

**Location of a percentile**

The following formula allows us to approximate the location of any percentile.

$$L_p = (n+1)\frac{p}{100}$$

where $L_p$ is the location of the $p^{th}$ percentile.

---

## Your turn

5, 5, 10, 12, 13, 14, 17, 19, 27, 28

Find $L_{25}$, $L_{50}$ and $L_{75}$.

## Interquartile Range (IQR)

- Measure of dispersion

$$IQR = Q_3 - Q_1$$

- Measure considers the spread in the middle 50% of the data.

- Not influenced by extreme values.


## Box and whisker Plot

::: columns
::: {.column width="50%"}

```{r, echo=FALSE}
library(ggplot2)
a <- c(1, 3, 4, 7, 8, 8, 9, 10, 11, 2, 5, 6, 4, 7, 3, 2, 2, 20, 20, 30)
matrix(a, nrow=2) |> knitr::kable()


```

```{r}
summary(a) 
```

**Let's draw the box and whisker plot**

:::
  
::: {.column width="50%"}



:::
:::
  

## Data

```{r}
# Generate sample data 
set.seed(25052020)
# normal
s1 <- rnorm(100)
s2 <- rbeta(100, 2, 1)
s3 <- rexp(100)
# mixture
s4 <- rnorm(100, rep(c(-1, 1), each = 50) * sqrt(0.9), sqrt(0.1))
four <- data.frame(
dist = factor(rep(c("s1", "s2", "s3", "s4"),
each = 100),
c("s1", "s2", "s3", "s4")),
vals = c(s1, s2, s3, s4)
)
df <- data.frame(s1=s1, s2=s2, s3=s3, s4=s4)


```

::: columns
::: {.column width="33%"}
```{r}
library(kableExtra)
df[1:33, ] |> knitr::kable() |> kable_styling(font_size = 18)
```
:::

::: {.column width="33%"}
```{r}
df[34:67, ] |> knitr::kable() |> kable_styling(font_size = 18)
```
:::

::: {.column width="34%"}
```{r}
df[68:100, ] |> knitr::kable() |> kable_styling(font_size = 18)
```
:::
:::

---

::: columns
::: {.column width="50%"}

**Histogram**

```{r, comment=NA, message=FALSE, echo=FALSE, fig.height=10}
library(ggplot2)
ggplot(four, aes(x = vals)) + 
geom_histogram(aes(fill = dist), colour = "black") + 
coord_cartesian(xlim = c(-3, 6)) + 
facet_wrap(~ dist) 
```

:::

::: {.column width="50%"}

**Box and whisker Plot**

```{r, comment=NA, message=FALSE, echo=FALSE, fig.height=10}
library(tidyverse)
df2 <- df |> pivot_longer(1:4, names_to = "group", values_to="value")
ggplot(df2, aes(x=group, y=value, fill=group)) + geom_boxplot()
```
:::
:::

## Your turn

Prices of Chocolate in Rupees (LKR):

50, 75, 100, 125, 150, 175, 200, 225, 250, 275


Prices of Chocolate in USD:

5, 7, 9, 12, 15, 18, 20, 23, 26, 0

Which group of chocolate prices exhibits the highest variation?

---

## Coefficient of variation

- Relative measure of variation

- It always expressed as a percentage rather than in terms of the units of the particular data.

- This is useful when comparing two or more sets of data that are measured in the different units.

$$CV = \frac{s}{\bar{x}}\times 100\%$$

# Measures of shape

---

## Skewness and the relationship of the mean, median and mode

In-class diagram: 



